---
title: "Confidence Intervals"
author: "[your name]"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
  html_document:
    toc: TRUE
---

```{r setup, message = FALSE}
knitr::opts_chunk$set(echo = TRUE)
# install.packages() if needed
library("infer")
library("moderndive")
library("patchwork")
library("tidyverse")
```

Source:  **Statistical Inference via Data Science:** *A Modern Dive into R and the Tidyverse*

* Chapter 8: Bootstrapping and Confidence Intervals
* [https://moderndive.com/8-confidence-intervals.html](https://moderndive.com/8-confidence-intervals.html)

# Setting: Pennies

Among pennies in circulation in 2019, what was the average year of minting?  We have a sample size of 50 pennies.

```{r}
# looking at the data set
pennies_sample
```

```{r}
# visualizing the the pennies
pennies_sample %>%
  ggplot(aes(x = year)) +
  geom_dotplot(binwidth = 1, color = "tan3", fill = "tan4") +
  labs(title = "Pennies Sample",
       subtitle = "observed in 2019",
       caption = "Source: Modern Dive",
       x = "year",
       y = "proportion")
```

```{r}
# visualizing the distribution of the pennies
p1 <- pennies_sample %>%
  ggplot(aes(x = year)) +
  geom_histogram(binwidth = 10, color = "tan3", fill = "tan4") +
  labs(title = "Pennies Sample",
       subtitle = "observed in 2019",
       caption = "Source: Modern Dive")

# display graph (in addition to storing the graph in a variable)
p1
```

```{r}
pennies_sample %>% summarize(xbar = mean(year))
```


# Resampling

Using the available sample of data to fabricate another sample is called *resampling*.

## Resampling Once

Suppose that we took the 50 pennies and resampled once while sampling *with replacement*.

```{r}
pennies_resampled_once <- pennies_sample %>%
  sample_n(size = 50, replace = TRUE)
```

```{r}
# visualizing the distribution of the pennies
p2 <- pennies_resampled_once %>%
  ggplot(aes(x = year)) +
  geom_histogram(binwidth = 10, color = "tan3", fill = "tan4") +
  labs(title = "Pennies Resampled Once",
       subtitle = "sampled with replacement",
       caption = "Source: Modern Dive")

# (using `patchwork` package to arrange plots side-by-side
p1 + p2
```

```{r}
pennies_resampled_once %>% summarize(xbar = mean(year))
```


## Resampled Many Times

Suppose now that we have each person in a 30-student discussion section repeat the act of drawing those 50 pennies with replacement.

```{r}
pennies_resampled_many <- pennies_sample %>%
  rep_sample_n(size = 50, replace = TRUE, reps = 30)
```

Now we have each virtual student report their mean year.

```{r}
pennies_resampled_many %>%
  group_by(replicate) %>%
  summarize(mean_year = mean(year))
```

```{r}
summary(pennies_sample$year)
```
```{r}
pennies_resampled_many %>%
  group_by(replicate) %>%
  mutate(mean_year = mean(year)) %>%
  ungroup() %>%
  select(replicate, mean_year) %>%
  distinct() %>%
  ggplot(aes(x = mean_year)) +
  geom_histogram(binwidth = 1, color = "tan3", fill = "tan4") +
  labs(title = "Resampling Results",
       subtitle = "N = 30 resamples",
       caption = "Source: Modern Dive")
```

Out of curiosity, let us push this process to $N = 1337$ resamples.

```{r}
pennies_resampled_means <- pennies_sample %>%
  rep_sample_n(size = 50, replace = TRUE, reps = 1337) %>%
  group_by(replicate) %>%
  mutate(mean_year = mean(year)) %>%
  ungroup() %>%
  select(replicate, mean_year) %>%
  distinct() 

pennies_resampled_means %>%
  ggplot(aes(x = mean_year)) +
  geom_histogram(binwidth = 1, color = "tan3", fill = "tan4") +
  labs(title = "Resampling Results",
       subtitle = "N = 1337 resamples",
       caption = "Source: Modern Dive")
```


# Confidence Intervals

## Toward Confidence Intervals

The standard deviation of a sampling distribution is called the *standard error*.

```{r}
xbar <- mean(pennies_resampled_means$mean_year)
SE   <- sd(pennies_resampled_means$mean_year)
```

We can build a 95% confidence interval by computing $\bar{x} \pm 1.96*SE$

```{r}
c(xbar - 1.96*SE, xbar + 1.96*SE)
```
```{r}
pennies_resampled_means %>%
  ggplot(aes(x = mean_year)) +
  geom_histogram(binwidth = 1, color = "tan3", fill = "tan4") +
  geom_vline(xintercept = c(xbar - 1.96*SE, xbar + 1.96*SE), color = "yellow", size = 2) +
  labs(title = "Resampling Results",
       subtitle = "N = 1337 resamples",
       caption = "Source: Modern Dive") +
  theme_minimal()
```

## Using the infer package

```{r}
pennies_sample %>%
  specify(response = year)
```

```{r}
pennies_sample %>%
  specify(response = year) %>%
  calculate(stat = "mean")
```

```{r}
pennies_sample %>%
  specify(response = year) %>%
  generate(reps = 1337, type = "bootstrap")
```

```{r}
bootstrap_distribution <- pennies_sample %>%
  specify(response = year) %>%
  generate(reps = 1337, type = "bootstrap") %>%
  calculate(stat = "mean")

# print
bootstrap_distribution
```

```{r}
visualise(bootstrap_distribution)
```

There are also wrappers in the `infer` package to extract the confidence interval

```{r}
bootstrap_distribution %>%
  get_confidence_interval(point_estimate = mean(bootstrap_distribution$stat), 
                          level = 0.95, type = "se")
```

Alternatively, we can use percentiles to build our confidence intervals.  This is useful when the data is not normally distributed.

```{r}
bootstrap_distribution %>%
  get_confidence_interval(level = 0.95, type = "percentile")
```

```{r}
SE_CI <- bootstrap_distribution %>%
  get_ci(point_estimate = mean(bootstrap_distribution$stat),
         level = 0.95, type = "se")

visualize(bootstrap_distribution) +
  shade_ci(endpoints = SE_CI, color = "#DAA900", fill = "#002856")
```

# Inference

The `bowl` data was literally a classroom bowl of red and white marbles

```{r}
bowl %>%
  ggplot(aes(x = color, fill = color)) +
  geom_bar(stat = "count", color = "black") +
  scale_fill_manual(values = c("red", "white")) +
  labs(title = "Bowl of Marbles",
       subtitle = "population is known",
       caption = "Source: Modern Dive")
```

where we know the true proportion of red marbles.

```{r}
bowl %>%
  summarize(proportion_red = mean(color == "red"))
```


```{r}
CI_simulation <- function(confidence = 95, sample_size = 25, num_intervals = 10){
  # Constants
  alpha <- 1 - confidence/100
  n <- sample_size
  N <- num_intervals
  proportion_red <- 0.375 #true population proportion
  
  # vector allocation
  left <- rep(NA, N)
  right <- rep(NA, N)
  captured <- rep(NA, N)
  
  for(i in 1:N){
    this_sample <- sample(bowl$color, n, replace = TRUE)
    phat <- mean(this_sample == "red") #sample proportion
    
    #margin of error
    E <- qnorm(1 - alpha/2)*sqrt( phat*(1-phat)/n)
    
    #this confidence interval
    left[i] <- phat - E
    right[i] <- phat + E
    
    #did the confidence interval capture the true proportion?
    captured[i] <- ifelse(left[i] <= proportion_red & right[i] >= proportion_red, TRUE, FALSE)
  }
  
  # graph
  df <- data.frame(left, right, captured)
  ggplot(df, aes(x = left, y = 1:N)) +
    geom_vline(xintercept = proportion_red, color = "black") +
    geom_segment(aes(x = left, y = 1:N, 
                     xend = right, yend = 1:N,
                     color = captured)) +
    labs(title = "Simulation of bowl samples",
         subtitle = paste0("alpha = ", alpha, ", n = ", n),
         caption = "Bio 175", 
         x = "proportion red",
         y = "iteration") +
    theme_minimal()
}
```

```{r}
CI_simulation(confidence = 95, sample_size = 25, num_intervals = 100)
```

```{r}
p1 <- CI_simulation(80, 25, 100) + theme(legend.position = "none")
p2 <- CI_simulation(95, 25, 100) + theme(legend.position = "none")
p3 <- CI_simulation(99, 25, 100) + theme(legend.position = "none")

p1 + p2 + p3
```

```{r}
p4 <- CI_simulation(95, 25, 100) + theme(legend.position = "none")
p5 <- CI_simulation(95, 100, 100) + theme(legend.position = "none")
p6 <- CI_simulation(95, 400, 100) + theme(legend.position = "none")

p4 + p5 + p6
```

